{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-05-04 paper.tex
edits

\begin{small}
\noindent
{\em Manuscript Title:} {\tt APFELgrid}: a fast interface to hadronic observables for
  fits of parton densities                                      \\
{\em Authors:} V.~Bertone, S.~Carrazza, N.P.~Hartland                                               \\
{\em Program Title:} {\tt APFELgrid}                                          \\
{\em Journal Reference:}                                      \\
  %Leave blank, supplied by Elsevier.
{\em Catalogue identifier:}                                   \\
  %Leave blank, supplied by Elsevier.
{\em Licensing provisions:}   MIT license                                \\
{\em Programming language:}  C++                                 \\
{\em Computer:} PC/Mac                                               \\
  %Computer(s) for which program has been designed.
{\em Operating system:} MacOS/Linux                                       \\
  %Operating system(s) for which program has been designed.
{\em RAM:} varying                                              \\
  %RAM in bytes required to execute program with typical data.
{\em Keywords:} QCD, PDF\\
  % Please give some freely chosen keywords that we can use in a
  % cumulative keyword index.
{\em Classification:}  11.1 General, High Energy Physics and Computing                                       \\
  %Classify using CPC Program Library Subject Index, see (
  % http://cpc.cs.qub.ac.uk/subjectIndex/SUBJECT_index.html)
  %e.g. 4.4 Feynman diagrams, 5 Computer Algebra.
{\em External routines/libraries:}  {\tt APPLgrid}, {\tt APFEL}                          \\
  % Fill in if necessary, otherwise leave out.
{\em Nature of problem:}\\
 Fast computation of hadronic observables under the variation of parton distribution functions.
   \\
{\em Solution method:}\\
  Combination of interpolated weight grids from {\tt APPLgrid} and evolution factors from {\tt APFEL} into efficient {\tt FastKernel} tables.\\
{\em Running time:} varying\\
   \\
\end{small}

@@

This interpolation strategy is implemented in the {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij} projects. For the computation of the
hard cross-sections, these packages rely on external codes to which
they are interfaced by means of a suite of functions allowing for the
filling of PDF- and $\alpha_s$-independent look-up tables of
cross-section weights. Codes such as {\tt MCFM}~\cite{Campbell:2010ff}
and {\tt NLOJet++}~\cite{Nagy:2003tz} have been interfaced directly to
{\tt APPLgrid}/{\tt FastNLO} and more recently dedicated interfaces to
automated general-purpose Monte Carlo (MC) event generators have been
developed. The {\tt aMCfast}~\cite{Bertone:2014zva} and {\tt
  MCgrid}~\cite{DelDebbio:2013kxa} codes have been specifically
designed to extract the relevant information from the {\tt
  MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators respectively.

While these tools have proven to be extremely useful in the extraction
of parton densities, the volume of experimental data made available by
LHC collaborations for use in PDF fits is already stretching the
capabilities of typical fitting technology. Presently a typical global
PDF fit includes thousands of hadronic data points for which
predictions have to be computed thousands of times during the
minimisation process. As a consequence the ``standard'' tools, $i.e.$
{\tt APPLgrid} and {\tt FastNLO}, might not be fast enough to meet the
needs of modern PDF fitters. For this reason a high-performance tool
tailored specifically for PDF analyses is becoming increasingly
important.

@@

Hadron collider observables are typically computed in QCD by a double
convolution of parton densities with a hard scattering
cross-section. Consider for example the calculation of a general
cross-section $pp\to X$ with a set of PDFs $\{f\}$:
\begin{equation}\label{eq:hadconv}
  \sigma_{pp\to X} =
  \sum_{s}\sum_{p} \int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\alpha_s^{p+p_{\rm LO}}(Q^2) \,F^{(s)}(x_1,x_2, Q^2)\,.
\end{equation}
where $Q^2$ is the typical hard scale of the process, the index
$s$ sums over the active partonic subprocesses in the calculation, $p_{\rm LO}$ is the leading-order power of $\alpha_s$ for the process
and $\hat{\sigma}^{(p)(s)}$ is the N$^p$LO cross-section for the partonic subprocess scattering $s\to X$. $F^{(s)}$ represents the 
subprocess parton density:
\begin{equation}\label{eq:APPLsubproc}
  F^{(s)}(x_1,x_2, Q^2) =\sum_{i,j} C^{(s)}_{ij} 
  f_i(x_{1},Q^2)f_j(x_{2},Q^2)\,,
\end{equation}
where the $C^{(s)}_{ij}$ matrix enumerates the combinations of PDFs
that contribute to the $s$-th subprocess.  The central observation of
tools such as {\tt APPLgrid} and {\tt FastNLO} is that the PDF and
$\alpha_S$ dependence may be factorised out of the convolution via
expansion over a set of interpolating functions, spanning $Q^2$ and
the two values of parton-$x$. For example one may represent the
subprocess PDFs and $\alpha_S$ in terms of the Lagrange basis
polynomials $\mathcal{I}_\tau(Q^2)$, $\mathcal{I}_\alpha(x_1)$ and
$\mathcal{I}_\beta(x_2)$ as:
\begin{equation}\label{eq:interpolation}
\begin{array}{c}
\displaystyle \alpha_s^{p+p_{\rm LO}}(Q^2) \,F^{(s)}(x_1,x_2, Q^2)
  =\\
\\
\displaystyle \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
  \,F^{(s)}_{\alpha\beta , \tau}
  \,\mathcal{I}_\tau(Q^2)\,\mathcal{I}_\alpha(x_1)
  \,\mathcal{I}_\beta(x_2)
\end{array}
\end{equation}
where we use the shorthand
$F^{(s)}_{\alpha\beta ,\tau} = F^{(s)}(x_\alpha, x_\beta,Q_\tau^2)$.
Using these expressions in the double-convolution of
Eq.~(\ref{eq:hadconv}) one can finally obtain an expression for the
desired cross-section which depends upon the subprocess PDFs only
though a simple product:

\begin{equation} \label{eq:applconv}
  \sigma_{pp\to X} = \sum_p^{N_\alpha} \sum_{s}^{N_{\mathrm{sub}}} \sum_{\alpha,\beta,\tau} 
  \alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \, F_{\alpha\beta,\tau}^{(s)}\,.
\end{equation}
where
\begin{equation}
  W_{\alpha\beta,\tau}^{(p)(s)} = \mathcal{I}_\tau(Q^2)\int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\mathcal{I}_\alpha(x_1)
  \,\mathcal{I}_\beta(x_2)\,,
\end{equation}
consists of the convolution of the hard cross-section with the
interpolating polynomials. This information may be stored in a
precomputed look-up table.  The final expression for the cross-section
in Eq.~(\ref{eq:applconv}) therefore being considerably more
computationally efficient inside a fit than the direct performance of
the double-convolution.

@@

A number of tools, $e.g.$ {\tt
  APFEL}~\cite{Bertone:2013vaa,Carrazza:2014gfa,Bertone:2015cwa,Bertone:2015lqa,Carrazza:2015dea,Bertone:2015gba}
or {\tt HOPPET}~\cite{Salam:2008qg}, are available which perform PDF
evolution via an analogous procedure. {\tt APFEL} in particular, is
able to provide the so-called \textit{evolution operator}. The
evolution operator, is computed assuming that the DGLAP evolution can
be encoded in a single operator whose action on the intial scale PDFs
proceduces the final scale PDFs. Taking $Q_0$ as an initial scale and
$Q_\tau$ as a final scale, on the $x$-space grid one can write:
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where the operator $A$ is the solution of the DGLAP evolution equation
with baundary condition:
\begin{equation}
A^\tau_{\alpha\beta, ik} =
\delta_{\alpha\beta}\delta_{ik}\qquad\mbox{for}\qquad Q_\tau = Q_0\,.
\end{equation}

Given the evolution operator, we may replace the (general-scale) PDFs
used in the subprocess parton density Eq.~(\ref{eq:APPLsubproc}) with
their equivalent expressions c.f Eq.~(\ref{eq:fastPDFfinal_recalled})
like so:
\begin{equation}\label{eq:FK1}
\begin{array}{rcl}
F^{(s)}_{\alpha\beta,\tau} &=&  \displaystyle \sum_{i,j} \sum_{k,l}
                               \sum_{\delta,\gamma} C^{(s)}_{ij}
                               \left[  A^\tau_{\alpha\delta ik}\;
                               f_k(x_\delta,Q^2_0) A^\tau_{\beta\gamma
                               jl}\; f_l(x_\gamma,Q^2_0) \right]\;\;\;
  \\
\\
&=& \displaystyle \sum_{k,l}\sum_{\delta,\gamma}
\widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,,
\end{array}
\end{equation}
where the object:
\begin{equation}
  \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta} =
  \sum_{i,j} C^{(s)}_{ij} A^\tau_{\alpha\delta ik}
  A^\tau_{\beta\gamma jl}
\end{equation}
combines the operations of subprocess density construction and PDF
evolution. Going further it is possible to then use the expression for
subprocess parton densities in Eq.~(\ref{eq:FK1}) in the full
cross-section calculation:
\begin{equation}
\sigma_{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma}\sum_p
\sum_{s} \sum_{\alpha,\beta}
\sum_{\tau} 
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,.
\end{equation}
Performing some further contractions it is possible to obtain an
extremely compact expression for the calculation of the cross section
in question, in terms of only the initial-scale PDFs and summing only
over the initial scale interpolating $x$-grid and the incoming parton
flavors:
\begin{equation}\label{eq:FK}
  \sigma _{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma} 
  \widetilde{W}_{kl,\delta\gamma} \,f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0),
\end{equation}
where the object:
\begin{equation}\label{eq:FKTable}
  \widetilde{W}_{kl,\delta\gamma} = \sum_p^{N_\alpha} \sum_{s}^{N_{\mathrm{sub}}} \sum_{\alpha,\beta} \sum_{\tau}
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)  W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
\end{equation}
is referred to here as an {\tt FK} table, and combines the information
stored in {\tt APPLgrid}-style interpolated weight grids with
analogously interpolated DGLAP evolution operators. This combination
enables for a maximally computationally-efficient expression for the
calculation of observables at hadron colliders under PDF variation.

@@

The {\tt FK} product of Eq.~(\ref{eq:FK}) differs with respect to the
product in Eq.~(\ref{eq:applconv}) in several notable ways. Firstly
the typical {\tt APPLgrid} or {\tt FastNLO} products require PDFs at a
general scale $Q^2$, which implies performing PDF evolution {\it e.g.}
Eq.~(\ref{eq:fastPDFfinal_recalled}) for every variation of the PDFs
at the input scale during the fit. However in the {\tt FK} convolution
this evolution is pre-cached at the stage of {\tt FK} table
generation, requiring only initial-scale PDFs as an input at fitting
time. This pre-caching of the evolution also removes the need for
performing the sums over hard scale and perturbative order during the
fit, further reducing the number of operations required.  The {\tt FK}
convolution also benefits from the typically reduced number of active
partonic modes at the fitting scale, with the sum over flavours in
Eq.~(\ref{eq:FK}) being limited to those directly used in the
fit. Having reduced the calculation to such a simple form, it is also
possible to straightforwardly apply standard and widely-available
computational tools such as multi-threading through e.g OpenMP or
Single Instruction Multiple Data (SIMD) operations e.g SSE or AVX to
further reduce computational cost.

@@

In Fig.~\ref{fig:timings} we compare the average time taken per
datapoint for the {\tt FK} and {\tt APPLgrid} calculations, for all of
the 52 tables. We show timings for the {\tt FK} calculation in four
different configurations: AVX-OpenMP 2x (2 CPU cores), AVX, SSE3 and
standard double precision dot product. Due to the inherent structural
differences between the {\tt FK} and {\tt APPLgrid} procedures,
results from the {\tt FK} calculation are systematically faster than
those from {\tt APPLgrids}.  In particular, when comparing {\tt FK}
AVX to {\tt APPLgrid} timings we obtain minimally a factor of ten
improvement in speed for electroweak vector boson production and a
maximum factor of 2000 improvement in predictions for inclusive jet
data. Across all processes and kinematic regions we observe
significant performance improvements from using the {\tt FK}
calculation even in the maximally compatible, non-SIMD case.

While sheer computational speed is typically the primary consideration
when computing observables in a PDF fit, other factors such as table
size in the filesystem and memory, along with the computational cost
of pre-computing {\tt FK} tables must be considered. Indeed, the
computation of the {\tt FK} table in Eq.~(\ref{eq:FKTable}) requires a
great deal of operations which can be time consuming, particuarly in
the case of source {\tt APPLgrids} with very high interpolation
precision.

In Fig.~\ref{fig:performance} we examine the {\tt FK} table generation
time with {\tt APFELgrid}, {\tt FK} table file size and memory usage
of the {\tt FK} tables arising from the same source {\tt APPLgrids} as
discussed in Fig.~\ref{fig:timings}.  When examining the table
generation time per point, we observe timings from a few milliseconds
to 3.5 minutes per point, with differences arising from the varying
grid densities used in the input {\tt APPLgrid} files.  In terms of
the grid size on disk, {\tt FK} tables are typically larger than their
corresponding {\tt APPLgrids}, primarily as the {\tt FK} file format
is encoding in plain text for compatibility whereas {\tt APPLgrids}
are expressed in binary as {\tt ROOT} files. However when measuring
the in-memory resident set size used by the both convolution
procedures, the amount of memory required by {\tt FK} tables is
systematically less than {\tt APPLgrid} for all processes considered
here by at least 75\%.

@@

The substantial speed improvement of {\tt FK} tables with respect to
{\tt APPLgrid} in association with the reduction in memory footprint,
makes the {\tt APFELgrid} code a valuable tool for modern PDF fits
including large collider datasets and thus rely heavily upon fast
calculations.

>>>>

\begin{small}
\noindent
{\em Manuscript Title:} {\tt APFELgrid}: a fast interface to hadronic observables for
  fits of parton densities                                      \\
{\em Authors:} V.~Bertone, S.~Carrazza, N.P.~Hartland                                               \\
{\em Program Title:} {\tt APFELgrid}                                          \\
{\em Journal Reference:}                                      \\
  %Leave blank, supplied by Elsevier.
{\em Catalogue identifier:}                                   \\
  %Leave blank, supplied by Elsevier.
{\em Licensing provisions:}   MIT license                                \\
{\em Programming language:}  C++                                 \\
{\em Computer:} PC/Mac                                               \\
  %Computer(s) for which program has been designed.
{\em Operating system:} MacOS/Linux                                       \\
  %Operating system(s) for which program has been designed.
{\em RAM:} varying                                              \\
  %RAM in bytes required to execute program with typical data.
{\em Keywords:} QCD, PDF\\
  % Please give some freely chosen keywords that we can use in a
  % cumulative keyword index.
{\em Classification:}  11.1 General, High Energy Physics and Computing                                       \\
  %Classify using CPC Program Library Subject Index, see (
  % http://cpc.cs.qub.ac.uk/subjectIndex/SUBJECT_index.html)
  %e.g. 4.4 Feynman diagrams, 5 Computer Algebra.
{\em External routines/libraries:}  {\tt APPLgrid}, {\tt APFEL}                          \\
  % Fill in if necessary, otherwise leave out.
{\em Nature of problem:}\\
 Fast computation of hadronic observables under the variation of parton distribution functions.
   \\
{\em Solution method:}\\
  Combination of interpolated weight grids from {\tt APPLgrid} files and evolution factors from {\tt APFEL} into efficient {\tt FastKernel} tables.\\
{\em Running time:} varying\\
   \\
\end{small}

@@

This interpolation strategy is implemented in the {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij} projects. For the computation of the
hard cross-sections, these packages rely on external codes to which
they are interfaced by means of a suite of functions allowing for the
filling of PDF- and $\alpha_s$-independent look-up tables of
cross-section weights. Codes such as {\tt MCFM}~\cite{Campbell:2010ff}
and {\tt NLOJet++}~\cite{Nagy:2003tz} have been interfaced directly to
{\tt APPLgrid}/{\tt FastNLO} and more recently de\-di\-ca\-ted
interfaces to automated general-purpose Monte Carlo (MC) event
generators have been developed. The {\tt
  aMCfast}~\cite{Bertone:2014zva} and {\tt
  MCgrid}~\cite{DelDebbio:2013kxa} codes have been specifically
designed to extract the relevant information from the {\tt
  MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators respectively.

While these tools have proven to be extremely useful in the extraction
of parton densities, the volume of experimental data made available by
LHC collaborations for use in PDF fits is already stretching the
capabilities of the typical fitting technology. Presently a typical
global PDF fit includes thousands of hadronic data points for which
predictions have to be computed thousands of times during the
minimisation process. As a consequence the ``standard'' tools, $i.e.$
{\tt APPLgrid} and {\tt FastNLO}, might not be fast enough to meet the
needs of modern PDF fitters. For this reason a high-performance tool
tailored specifically for PDF analyses is becoming increasingly
important.

@@

Hadron collider observables are typically computed in QCD by a double
convolution of parton densities with a hard scattering
cross-section. Consider for example the calculation of a general
cross-section $pp\to X$ with a set of PDFs $\{f\}$:
\begin{equation}\label{eq:hadconv}
  \sigma_{pp\to X} =
  \sum_{s}\sum_{p} \int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\alpha_s^{p+p_{\rm LO}}(Q^2) \,F^{(s)}(x_1,x_2, Q^2)\,.
\end{equation}
where $Q^2$ is the typical hard scale of the process, the index $s$
sums over the active partonic subprocesses in the calculation, $p_{\rm
  LO}$ is the leading-order power of $\alpha_s$ for the process and
$\hat{\sigma}^{(p)(s)}$ is the N$^p$LO contribution to the
cross-section for the partonic subprocess scattering $s\to
X$. $F^{(s)}$ represents the subprocess parton density:
\begin{equation}\label{eq:APPLsubproc}
  F^{(s)}(x_1,x_2, Q^2) =\sum_{i,j} C^{(s)}_{ij} 
  f_i(x_{1},Q^2)f_j(x_{2},Q^2)\,,
\end{equation}
where the $C^{(s)}_{ij}$ matrix enumerates the combinations of PDFs
that contribute to the $s$-th subprocess.  The central observation of
tools such as {\tt APPLgrid} and {\tt FastNLO} is that the PDF and
$\alpha_S$ dependence may be factorised out of the convolution via
expansion over a set of interpolating functions, spanning $Q^2$ and
the two values of parton-$x$. For example one may represent the
subprocess PDFs and $\alpha_S$ in terms of the Lagrange basis
polynomials $\mathcal{I}_\tau(Q^2)$, $\mathcal{I}_\alpha(x_1)$ and
$\mathcal{I}_\beta(x_2)$ as:
\begin{equation}\label{eq:interpolation}
\begin{array}{c}
\displaystyle \alpha_s^{p+p_{\rm LO}}(Q^2) \,F^{(s)}(x_1,x_2, Q^2)
  =\\
\\
\displaystyle \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
  \,F^{(s)}_{\alpha\beta , \tau}
  \,\mathcal{I}_\tau(Q^2)\,\mathcal{I}_\alpha(x_1)
  \,\mathcal{I}_\beta(x_2)
\end{array}
\end{equation}
where we use the shorthand $F^{(s)}_{\alpha\beta ,\tau} =
F^{(s)}(x_\alpha, x_\beta,Q_\tau^2)$.  Using these expressions in the
double convolution of Eq.~(\ref{eq:hadconv}) one can finally obtain an
expression for the desired cross-section which depends upon the
subprocess PDFs only through a simple product:

\begin{equation} \label{eq:applconv}
  \sigma_{pp\to X} = \sum_p \sum_{s}\sum_{\alpha,\beta,\tau} 
  \alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \, F_{\alpha\beta,\tau}^{(s)}\,.
\end{equation}
where:
\begin{equation}
  W_{\alpha\beta,\tau}^{(p)(s)} = \mathcal{I}_\tau(Q^2)\int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\mathcal{I}_\alpha(x_1)
  \,\mathcal{I}_\beta(x_2)\,,
\end{equation}
consists of the convolution of the hard cross-section with the
interpolating polynomials. This information may be stored in a
precomputed look-up table.  The final expression for the cross-section
in Eq.~(\ref{eq:applconv}) therefore being considerably more
computationally efficient inside a fit than the direct performance of
the double convolution.

@@

A number of tools, $e.g.$ {\tt
  APFEL}~\cite{Bertone:2013vaa,Carrazza:2014gfa,Bertone:2015cwa,Bertone:2015lqa,Carrazza:2015dea,Bertone:2015gba},
{\tt HOPPET}~\cite{Salam:2008qg} and {\tt QCDNUM}~\cite{Botje:2010ay},
are available which perform PDF evolution via an analogous
procedure. {\tt APFEL} in particular, is able to provide the so-called
\textit{evolution operator}. The evolution operator, is computed
assuming that the DGLAP evolution can be encoded in a single operator
whose action on the intial scale PDFs proceduces the final scale
PDFs. Taking $Q_0$ as an initial scale and $Q_\tau$ as a final scale,
on the $x$-space grid one can write:
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where the operator $A$ is the solution of the DGLAP evolution equation
with baundary condition:
\begin{equation}
A^\tau_{\alpha\beta, ik} =
\delta_{\alpha\beta}\delta_{ik}\qquad\mbox{for}\qquad Q_\tau = Q_0\,.
\end{equation}

Given the evolution operator defined in
Eq.~(\ref{eq:fastPDFfinal_recalled}), we may replace the
(general-scale) PDFs used in the subprocess parton density
Eq.~(\ref{eq:APPLsubproc}) with their equivalent expressions like so:
\begin{equation}\label{eq:FK1}
\begin{array}{rcl}
F^{(s)}_{\alpha\beta,\tau} &=&  \displaystyle \sum_{i,j} \sum_{k,l}
                               \sum_{\delta,\gamma} C^{(s)}_{ij}
                               \left[  A^\tau_{\alpha\delta ik}\;
                               f_k(x_\delta,Q^2_0) A^\tau_{\beta\gamma
                               jl}\; f_l(x_\gamma,Q^2_0) \right]\;\;\;
  \\
\\
&=& \displaystyle \sum_{k,l}\sum_{\delta,\gamma}
\widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,,
\end{array}
\end{equation}
where the object:
\begin{equation}
  \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta} =
  \sum_{i,j} C^{(s)}_{ij} A^\tau_{\alpha\delta ik}
  A^\tau_{\beta\gamma jl}\,,
\end{equation}
combines the operations of subprocess density construction and PDF
evolution. Going further it is possible to then use the expression for
subprocess parton densities in Eq.~(\ref{eq:FK1}) in the full
cross-section calculation:
\begin{equation}
\sigma_{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma}\sum_p
\sum_{s} \sum_{\alpha,\beta}
\sum_{\tau} 
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,.
\end{equation}
Performing some further contractions it is possible to obtain an
extremely compact expression for the calculation of the cross section
in question, in terms of only the initial-scale PDFs and summing only
over the initial scale interpolating $x$-grid and the incoming parton
flavors:
\begin{equation}\label{eq:FK}
  \sigma _{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma} 
  \widetilde{W}_{kl,\delta\gamma} \,f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0),
\end{equation}
where the object:
\begin{equation}\label{eq:FKTable}
  \widetilde{W}_{kl,\delta\gamma} = \sum_p\sum_{s}\sum_{\alpha,\beta} \sum_{\tau}
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)  W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
\end{equation}
is referred to here as an {\tt FK} table, and combines the information
stored in {\tt APPLgrid}-style interpolated weight grids with
analogously interpolated DGLAP evolution operators. This combination
enables for a maximally computationally-efficient expression for the
calculation of observables at hadron colliders under PDF variation.

@@

The {\tt FK} product of Eq.~(\ref{eq:FK}) differs with respect to the
product in Eq.~(\ref{eq:applconv}) in several notable ways. Firstly
the typical {\tt APPLgrid} or {\tt FastNLO} products require PDFs at a
general scale $Q^2$, which implies performing PDF evolution {\it e.g.}
Eq.~(\ref{eq:fastPDFfinal_recalled}) for every variation of the PDFs
at the input scale during the fit. However in the {\tt FK} convolution
this evolution is pre-cached at the stage of {\tt FK} table
generation, requiring only initial-scale PDFs as an input at fitting
time. This pre-caching of the evolution also removes the need for
performing the sums over hard scale and perturbative order during the
fit, further reducing the number of operations required. The {\tt FK}
convolution also benefits from the typically reduced number of active
partonic modes at the fitting scale, with the sum over flavors in
Eq.~(\ref{eq:FK}) being limited to those directly used in the
fit. Having reduced the calculation to such a simple form, it is also
possible to straightforwardly apply standard and widely-available
computational tools such as multi-threading through e.g OpenMP or
Single Instruction Multiple Data (SIMD) operations e.g SSE or AVX to
further reduce computational cost.

@@

In Fig.~\ref{fig:timings} we compare the average time taken per
datapoint for the {\tt FK} and {\tt APPLgrid} calculations, for all of
the 52 tables. We show timings for the {\tt FK} calculation in four
different configurations: AVX-OpenMP 2x (2 CPU cores), AVX, SSE3 and
standard double precision dot product. Due to the inherent structural
differences between the {\tt FK} and {\tt APPLgrid} procedures,
results from the {\tt FK} calculation are systematically faster than
those from {\tt APPLgrid}. In particular, when comparing {\tt
  FK} AVX to {\tt APPLgrid} timings we obtain minimally a factor of
ten improvement in speed for electroweak vector boson production and a
maximum factor of 2000 improvement in predictions for inclusive jet
data. Across all processes and kinematic regions we observe
significant performance improvements from using the {\tt FK}
calculation even in the maximally compatible, non-SIMD case.

While sheer computational speed is typically the primary consideration
when com\-pu\-ting observables in a PDF fit, other factors such as
table size in the filesystem and memory, along with the computational
cost of pre-computing {\tt FK} tables must be considered. Indeed, the
computation of the {\tt FK} table in Eq.~(\ref{eq:FKTable}) requires a
great deal of operations which can be time consuming, particuarly in
the case of source {\tt APPLgrid} files with very high interpolation
precision.

In Fig.~\ref{fig:performance} we examine the {\tt FK} table generation
time with {\tt APFELgrid}, {\tt FK} table file size and memory usage
of the {\tt FK} tables arising from the same source {\tt APPLgrid}
files as discussed in Fig.~\ref{fig:timings}.  When examining the
table generation time per point, we observe timings from a few
milliseconds to 3.5 minutes per point, with differences arising from
the varying grid densities used in the input {\tt APPLgrid} files.  In
terms of the grid size on disk, {\tt FK} tables are typically larger
than their corresponding {\tt APPLgrid} files, primarily as the {\tt
  FK} file format is encoded in plain text for compatibility whereas
{\tt APPLgrid} files are expressed in binary as {\tt ROOT}
files. However when measuring the in-memory resident set size used by
the both convolution procedures, the amount of memory required by {\tt
  FK} tables is systematically less than {\tt APPLgrid} files for all
processes considered here by at least 75\%.

@@

The substantial speed improvement of {\tt FK} tables with respect to
{\tt APPLgrid} in association with the reduction in memory footprint,
makes the {\tt APFELgrid} code a valuable tool for modern PDF fits
including large collider datasets and that thus rely heavily upon fast
calculations.
